SHARP - II: Revealing a bias in observational measurements of dark matter substructure with gravitational lens flux ratios

### 2015-11-11 papers/B1555_EdgeOn/Resubmission/referee_response.txt

We thank the referee for his/her comments about the paper.  We have
responded to the comments below, and modified the text accordingly.
Overall, we have toned down our somewhat sweeping claims.

> Comments to the Author: This paper presents an interesting new
> observational result for the strong gravitational lens B1555+375,
> along with an abundance of poorly supported conjecture.  The
> observations of B1555+375 should be published, but the authors should
> either remove their speculations, or (more usefully) provide evidence
> in support of them.
> 
> CLASS B1555+375 is one of the "poster child" gravitational lens
> systems.  It is famous for its prototypical flux ratio anomaly, which
> has been cited as evidence for dark matter substructure in this
> system.  The new result presented in this paper is near-infrared
> imaging with Keck and HST which reveals a linear structure in the lens
> galaxy emission consistent with an edge-on thin disk.  Most
> intriguingly, this structure passes near the anomalous images, and
> could explain the observed flux anomaly without the need to invoke
> dark matter substructure.  This would be an important result, and
> hence should be published.
> 
> However, the discussion in the paper lacks important quantitative
> details.  For example, the paper does not describe the total
> luminosity of the disky component, compared to the spheroidal
> component of the lens galaxy.  This is important because the lens
> model discussed in the paper requires the disk component to comprise ~
> 10% of the mass within the Einstein radius, and it is unclear if that
> large mass fraction is consistent with the photometry shown in Figure
> 1.
>

RESPONSE: The referee is correct to point out that we did not quantify
the bulge/disk ratio in the lensing galaxy.  We have performed a
bulge+disk fit to the lens-galaxy surface brightness distribution.
Our fit returns a bulge component that is consistent with zero
emission.  [DO WE WANT TO MAKE M/L RATIO CALCULATIONS??].  We have
added text to the paper describing this result.  We note that the two
components in our lens model can be thought of as the edge-on disk and
the dark-matter halo of the lensing galaxy.

> More importantly, the large disk component in the lens model produces
> a significant distortion to the usually ellipsoidal critical curves
> found in strong lenses, as shown in Fig 3.  It has long been known
> that such departures from ellipsoidal density distributions are
> tightly constrained by extended emission from the source galaxies,
> e.g. arcs and Einstein rings that span a large range of azimuthal
> angle.  The imaging shown in Fig 2 exhibits faint extended arclets
> from the source galaxy that could constrain disky components in the
> lens mass distribution, but the lens modeling in the paper neglects to
> fit the extended emission, and instead only models the positions and
> fluxes of the pointlike radio quasar.  The signal-to-noise of the
> extended emission may be too low to provide meaningful constraints,
> but given that the authors have this data in hand, and given that the
> authors have considerable experience modeling this type of extended
> emission in lens systems, they should attempt to fit the observed
> arclets to see if a 10% mass fraction in an off-center thin disk is
> consistent with the observed arclets.

RESPONSE: The referee is correct in pointing out that the modelling of
the extended images would provide tighter constraints on the proposed
model. Unfortunately, in the case of B1555 IR imaging the disk is
responsible for a significant level of dust extinction that leave
images D and B unobservable, and even the arc near image A strongly
affected. We have, however, tried modelling this dataset by only
focusing on image C which is not affected by dust, and the results are
consistent with those found by modelling the point images positions
and fluxes. For the VLBI extended images the poor quality of the data
prevented us from modelling this dataset in a reliable manner.  
[ADD TEXT TO PAPER??]

> 
> The final section of the paper speculates that massive disc components
> may be prevalent in strong lenses, and that such discs may bias
> inferences on the abundance of dark matter substructure using strong
> lensing.  The authors fail to provide evidence supporting these
> speculations, even though they have the data to do so.  As noted
> above, strong lenses with extended source emission (arcs and rings)
> tightly constrain the presence of thin components in lens galaxies.
> In some cases with Einstein rings, disc components are excluded from
> contributing more than percent-level fractions of the mass density.
> It would therefore be highly surprising for a substantial fraction of
> strong lenses to contain hidden, massive edge-on discs that could
> distort the lensing so significantly, as the paper intimates.
> 
> Rather than wildly speculating, the authors could easily check the
> plausibility of this claim.  A number of the authors were members of
> the SLACS collaboration, whose dataset contains many lenses with
> extended emission from the source galaxies.  These systems are the
> obvious way to quantify the fraction of lens systems containing
> massive edge-on discs, since the arcs in many SLACS lenses span a wide
> range of azimuthal angle.  Once they have quantified the abundance of
> massive edge-on discs, the authors should then perform Monte Carlo
> calculations of quasar lenses to calculate the statistics of flux
> ratio anomalies produced by these discs.  Only then can they state
> whether hidden edge-on massive discs can bias observational
> measurements of dark matter substructure with gravitational lens flux
> ratios.

RESPONSE: In the SHARP sample, we find that 5 out of the 25 systems
that we imaged have clear disk components, indicating that this is not
a completely unexpected phenomenon.  While we agree with the referee
that it is important to quantify how common disks are in lensing
galaxies, we do not feel that the SLACS sample provides the
appropriate comparison.  This is because a significant portion of the
SHARP systems come from source-selected samples, such as CLASS, that
are agnostic about the nature of the lensing galaxy.  In contrast,
SLACS lenses were selected to be systems where the lensing galaxy was
known a priori to be a massive elliptical galaxy.  We would therefore
expect a lower incidence of disky systems in SLACS than in the SHARP
sample.  We agree completely with the referee that doing a
quantification of the statistics of massive disks and their effects on
flux-ratio anomalies is an extremely important task and, in fact, this
is one of the future projects that we are planning.  However, we feel
that such an extensive analysis is beyond the scope of this letter.
As such, we have modified the text to focus more on this individual
lens system and put more caveats on the possible generalization of the
results.

### 2015-11-12 papers/B1555_EdgeOn/B1555_letter.tex
Incorporate Matt's text into conclusions of paper

The best-fit model parameters and 68\% CL errors are listed in Table~\ref{tab:model} and the lens model is illustrated in Fig.~\ref{fig:model}. Table \ref{tab:results} also shows the comparison between the observed radio data and the model-predicted image positions and flux ratios. We find that by adding an elongated exponential disc to a smooth SIE model there is significant improvement in the predicted flux-ratios when compared to the results for the SIE-only model. 
% with a reduced $\chi^2 = 2.7$
The positions are all in agreement, and the flux-ratios are consistent at the $2\sigma$ level, with excellent agreement now obtained for the two merging images A and B. The remaining differences in fluxes may due to the real lensed source being slightly extended rather than the point source that we used in our model. In IR imaging there is no sign of bulge, we then interpret the mass in ExpDisc component as the baryonic mass from lens galaxy.The model requires that the ExpDisc component makes up about 10 per cent of the total mass within the Einstein radius (1.7~kpc for a lens redshift of $z_l = 1$), which is not unreasonable.  We note that there is a significant 0.036~arcsec offset between the SIE and ExpDisc components, which is likely due to an additional ellipticity in the model (e.g. from an external shear) that is being partly compensated for by the disc component. Again, more complex mass models need the full extent of the gravitational arc for constraints. While the gravitational arc in IR band suffers severe dust extinction, the compact lensed images from future high resolution radio imaging will be able to confirm this.

@@

Here, we find evidence for an edge-on disc in the flux-ratio anomalous gravitational lens B1555+375 that can also account for the required perturbation in the mass model. Such high-order deviations from a smooth mass distribution have already been suggested \citep{evans03,congdon05}, but this is the first time that a disc has been observed and modelled to account for a flux-ratio anomaly. However, it is important to realise that by including observationally motivated perturbations like luminous dwarf companions or edge-on discs in the mass models, we are still limited in our interpretation since we must assume that the anomalies are due exclusively to these additional components. For B1555+375, it may be possible to independently constrain the mass of the disc through kinematic measurements (most likely from mm-wavelength resolved emission line observations), which would test whether the disc makes a sufficient contribution to the mass model (e.g. as in the case of B1933+503; \citealt{suyu12}). Also, as we have shown here, B1555+375 shows evidence for an extended $\sim100$~mas gravitational arc at radio-wavelengths that crosses the edge-on disc. Further high resolution imaging with very long baseline interferometry will provide the observational constraints required to fully test whether the extended disc is sufficient to cause the anomaly. Nevertheless, before concluding that the flux-ratios in any given lens system are a result of substructure, it is critical to explore a range of smooth models that are informed by additional observations or to do a comprehensive analysis such as that conducted by \citet{Xu15}. Otherwise, the derived substructure fractions should be considered as upper limits.

Our analysis for B1555+375 shows that in order to derive a precise quantification of the substructure population and other properties from strongly lensed flux-ratio anomalies, it is crucial to have a careful estimation of non-substructure effects, such as galaxy discs and nearby massive satellites. 
%We find that in B1555+375, a lack of knowledge about the lens galaxy mass distribution, combined with the limited information provided by the position of compact lensed images, can lead to improper lens models and a misinterpretation of the origin of flux-ratio anomalies. 
While upcoming large-scale surveys with the Dark Energy Survey, Kilo-Degree Sky-VIKING, Square Kilometre Array, {\it Euclid}, and Large Synoptic Survey Telescope, are expected to deliver new large samples of gravitationally lensed quasars, in the near future the discovery of new systems using narrow-line quasar emission \citep{N14} will provide a new and invaluable statistical perspective on the current constraints based on flux-ratio anomalies. We stress, therefore, that one should be very careful when interpreting the nature of flux-ratio anomalies, as prior simple assumptions on the lens mass distribution may a significant source of bias. Finally, additional near-infrared AO imaging from SHARP shows that B1555+375 is not the only gravitational lens having a significant disc component detection in the lensing galaxy. 5 out of 25 SHARP lenses are found discs in AO imaging (Fassnacht et al., in prep.).  These gravitational lenses will be investigated in a future work, but the images alone strongly suggest that other systems may have similar links between flux-ratio anomalies and disc components.

>>>>

The best-fit model parameters and 68\% CL errors are listed in Table~\ref{tab:model} and the lens model is illustrated in Fig.~\ref{fig:model}. Table \ref{tab:results} also shows the comparison between the observed radio data and the model-predicted image positions and flux ratios. We find that by adding an elongated exponential disc to a smooth SIE model there is significant improvement in the predicted flux-ratios when compared to the results for the SIE-only model. 
% with a reduced $\chi^2 = 2.7$
The positions are all in agreement, and the flux-ratios are consistent at the $2\sigma$ level, with excellent agreement now obtained for the two merging images A and B. The remaining differences in fluxes may due to the real lensed source being slightly extended rather than the point source that we used in our model. {\bf In the IR imaging there is no sign of a bulge; a formal bulge plus disk fit to the lens galaxy surface brightness distribution gives a bulge component that is consistent with zero emission.  We therefore interpret the mass in the ExpDisc component as the baryonic mass of lens galaxy.} The model requires that the ExpDisc component makes up about 10 per cent of the total mass within the Einstein radius (1.7~kpc for a lens redshift of $z_l = 1$), which is not unreasonable.  We note that there is a significant 0.036~arcsec offset between the SIE and ExpDisc components, which is likely due to an additional ellipticity in the model (e.g. from an external shear) that is being partly compensated for by the disc component. Again, more complex mass models need the full extent of the gravitational arc for constraints. {\bf The extinction in the AO imaging is so severe that components B and D are not visible, and the arc that connects A and B appears to be truncated.  The low signal-to-noise ratio and the strong extinction prevent us from using modeling techniques that incorporate the extended NIR emission \citep[e.g.][]{V12}.  Future high resolution radio imaging is needed in order to place stronger constraints on the mass model.}

@@

{\bf Here, we find evidence for an edge-on disc in the flux-ratio
anomalous gravitational lens B1555+375 that can also account for
the required perturbation in the mass model. Such high-order deviations from a smooth mass distribution have already been suggested previously (Evans \& Witt 2003; Congdon \& Keeton 2005), but this is the first time that an observed disc has been explicitly modelled to account for a flux-ratio anomaly. However, it is important to realise that by including observationally-motivated perturbations to a `smooth' mass distribution, like luminous dwarf companions or edge-on discs, in the mass models,
we are still limited in our interpretation since we must assume that
the anomalies are due exclusively to these additional components.
For B1555+375, it may be possible to independently constrain the
mass of the disc through kinematic measurements (most likely from
mm-wavelength resolved emission line observations), which would
test whether the disc makes a sufficient contribution to the mass
model (e.g. as in the case of B1933+503; Suyu et al. 2012). Also, as
we have shown here, B1555+375 shows evidence for an extended $\sim$ 100 mas gravitational arc at radio wavelengths that crosses the
edge-on disc. Further high resolution imaging with very long baseline interferometry will provide the observational constraints required to fully test whether the disc is sufficient to cause the anomaly.}

{\bf Our analysis of B1555+375 demonstrates that the choice of the macromodel used to describe the `smooth' lens galaxy mass distribution, combined with
the limited information provided by the position of compact lensed
images, can lead to a misinterpretation of the origin of flux-ratio anomalies. Before concluding that the flux-ratio anomalies in any given lens system are the result of dark matter substructure, then, it is critical that other possible sources of significant curvature in the lensing potential (i.e., `sharp' mass features such as edge-on disks or even massive luminous satellites) are taken into account (e.g., Xu et al.~2015). This result is timely, as upcoming large-scale surveys (e.g., the Dark Energy Survey, Kilo-Degree Survey, Euclid, and the Large Synoptic Survey Telescope) will deliver orders of magnitude more gravitationally lensed quasars for which flux ratios can be measured using narrow lines (e.g., Nierenberg et al.~2014). Near-infrared AO imaging of other lenses in the SHARP sample shows that B1555+375 is not the only system having a significant disc component detection in the lensing galaxy; these gravitational lenses will be investigated in future work, but the images alone suggest that other systems may have similar links between flux-ratio anomalies and disc components. Finally, we note that massive edge-on discs are \textit{not} observed in some flux-ratio anomaly systems (and the presence of a disc-like component may not resolve the anomaly anyways; Evans \& Witt 2003), indicating that these systems are still useful for probing the dark matter substructure mass function.}

%Here, we find evidence for an edge-on disc in the flux-ratio anomalous gravitational lens B1555+375 that can also account for the required perturbation in the mass model. Such high-order deviations from a smooth mass distribution have already been suggested \citep{evans03,congdon05}, but this is the first time that a disc has been observed and modelled to account for a flux-ratio anomaly. However, it is important to realise that by including observationally motivated perturbations like luminous dwarf companions or edge-on discs in the mass models, we are still limited in our interpretation since we must assume that the anomalies are due exclusively to these additional components. For B1555+375, it may be possible to independently constrain the mass of the disc through kinematic measurements (most likely from mm-wavelength resolved emission line observations), which would test whether the disc makes a sufficient contribution to the mass model (e.g. as in the case of B1933+503; \citealt{suyu12}). Also, as we have shown here, B1555+375 shows evidence for an extended $\sim100$~mas gravitational arc at radio-wavelengths that crosses the edge-on disc. Further high resolution imaging with very long baseline interferometry will provide the observational constraints required to fully test whether the extended disc is sufficient to cause the anomaly. Nevertheless, before concluding that the flux-ratios in any given lens system are a result of substructure, it is critical to explore a range of smooth models that are informed by additional observations or to do a comprehensive analysis such as that conducted by \citet{Xu15}. Otherwise, the derived substructure fractions should be considered as upper limits.

%Our analysis for B1555+375 shows that in order to derive a precise quantification of the substructure population and other properties from strongly lensed flux-ratio anomalies, it is crucial to have a careful estimation of non-substructure effects, such as galaxy discs and nearby massive satellites. 
%We find that in B1555+375, a lack of knowledge about the lens galaxy mass distribution, combined with the limited information provided by the position of compact lensed images, can lead to improper lens models and a misinterpretation of the origin of flux-ratio anomalies. 
%While upcoming large-scale surveys with the Dark Energy Survey, Kilo-Degree Sky-VIKING, Square Kilometre Array, {\it Euclid}, and Large Synoptic Survey Telescope, are expected to deliver new large samples of gravitationally lensed quasars, in the near future the discovery of new systems using narrow-line quasar emission \citep{N14} will provide a new and invaluable statistical perspective on the current constraints based on flux-ratio anomalies. We stress, therefore, that one should be very careful when interpreting the nature of flux-ratio anomalies, as prior simple assumptions on the lens mass distribution may a significant source of bias. Finally, additional near-infrared AO imaging from SHARP shows that B1555+375 is not the only gravitational lens having a significant disc component detection in the lensing galaxy. 5 out of 25 SHARP lenses are found discs in AO imaging (Fassnacht et al., in prep.).  These gravitational lenses will be investigated in a future work, but the images alone strongly suggest that other systems may have similar links between flux-ratio anomalies and disc components.
